Combinatorial interpretations of holonomic recurrences

### 2021-07-19 article/main.tex
Finish Schroder without images

@@

>>>>

The generating function equation is then:

\[\left\{\begin{array}{rcl}
            \gf{T}(z) &=& \gf{T}_1(z) + \gf{T}_2(z) + z\\
            \gf{T}_1(z) &=& \gf{T}^2(z)\\
            \gf{T}_2(z) &=& \gf{T}(z)(\gf{T}_1(z) + \gf{T}_2(z))
\end{array}\right.\]

To be able to use Maple to generate the recurrence, we need to transform this system of equations to a single one. We can eliminate \(\gf{T}_1(z)\) first.

\[\left\{\begin{array}{rcl}
            \gf{T}(z) &=& \gf{T}^2(z) + \gf{T}_2(z) + z\\
            \gf{T}_2(z) &=& \gf{T}(z)(\gf{T}^2(z) + \gf{T}_2(z))
\end{array}\right.\]

\[\left\{\begin{array}{rcl}
            \gf{T}(z) &=& \gf{T}^2(z) + \gf{T}_2(z) + z\\
            \gf{T}_2(z)(1 - \gf{T}(z)) &=& \gf{T}^3(z)
\end{array}\right.\]

As we can not use division to get \(\gf{T}_2(z)\), let's multiply the first equation by \(1 - \gf{T}(z)\).

\[\gf{T}(z)(1 - \gf{T}(z)) = \gf{T}^2(z)(1 - \gf{T}(z)) + \gf{T}^3(z) + z(1 - \gf{T}(z))\]

It can be actually used now to generate the holonomic recurrence.

\begin{lstlisting}
> with(gfun):
> RootOf(T*(1 - T) = T^2*(1 - T) + T^3 + z*(1 - T),  T);
               2
    RootOf(2 _Z  + (-z - 1) _Z + z)

> algfuntoalgeq(%, T(z));
       2
    2 T  + (-z - 1) T + z

> algeqtodiffeq(%, T(z));
                               2            /d      \                    2
    {z - 1 + (-z + 3) T(z) + (z  - 6 z + 1) |-- T(z)|, T(0) = RootOf(2 _Z  - _Z)}
                                            \dz     /

> diffeqtorec(%, T(z), T(n));

    {(n - 1) T(n) + (-3 - 6 n) T(n + 1) + (n + 2) T(n + 2), ...}

> subs(n = n - 1, %);

    {(-2 + n) T(n - 1) + (3 - 6 n) T(n) + (n + 1) T(n + 1), ...}
\end{lstlisting}

We can also move \(\gf{T}(z)\) to the other side, ie.:

\[3 (2 n - 1) T(n) = (n - 2) T(n - 1) + (n + 1) T(n + 1)\]

This may seem counterintuitive, as by default we want to describe how to generate larger case from the smaller ones, but it will make the interpretation simpler.

Let's translate the recurrence to the bijective language then.
\begin{itemize}
    \item \((2 n - 1) T(n)\) describes well-weighted binary plane trees of size \(n\) (ie. with \(n\) leaves) with some node pointed (internal or leaf),
    \item \((n + 1) T(n + 1)\) describes well-weighted binary plane trees of size \(n - 1\) with some leaf pointed,
    \item \((n - 2) T(n - 1)\) describes well-weighted binary plane trees of size \(n - 1\) with some internal node pointed.
\end{itemize}

We will show the bijection between two sides of the recursion. We will transform each tree from \((2 n - 1) T(n)\) in three ways in order to achieve all trees from \((n + 1) T(n + 1)\). We will get some extra ones which are not well-weighted, but we will deal with them in a minute.

Let's define 3 transformations: \(L_1\), \(L_2\) and \(R_1\).

\todo{image of the transformations}

You can see that trees generated by \(L_1\) and \(R_1\) are always well-weighted. The problem appears when applying \(L_2\) when \(s\) is a leaf.

When \(s\) is a leaf, we will look at the \(s\)'s parent. It can fall into one of following cases:

\todo{image of the problematic cases}

where \(t'\) can be any tree and \(t''\) can not be a leaf.

When aplying \(L_2\) we achieve following cases respectively:

\todo{images}

In case 1. and 2. we can just swap the labels as follows:

\todo{images}

It is easy to show that it constructs bijection between \((n + 1) T(n + 1)\) and subset of \(\{L_1, L_2, R_1\} \times (2 n - 1) T(n)\) not falling into case 3.

To achive trees from \((n - 2) T(n - 1)\), we can simply take trees from case 3. and replace the father of the \(s\) with \(t''\) and point root of \(t''\). In this way, we will loose \(2\) leaves and obtain each tree from \((n - 2) T(n - 1)\).

\todo{image}

@@

\subsection{Introduction}

As you could see in the examples above, most of the solutions base on the simple analysis of number of laves, internal nodes and other similar properties. In this section, we've gathered some methods that may be helpful when trying to analyze the problem of interpreting combinatorically some holonomic recurrence.
